---
output: html_document
---

## NBA Playoff Prediction

## File: main.rmd

This file is used for data cleaning and model building to predict the NBA 2023-2024 playoff result from round1 to final.

```{r}
# import packages
install.packages("randomForest")
library(randomForest)
```

```{r}
# Import functions from function.Rmd
knitr::read_chunk("function.Rmd")
```

## Part 1. Data Cleaning

### Part 1.1 Import dataset

-   Import team_game_data.csv as team_data.

```{r}
team_data <- read.csv("data/team_game_data.csv")
```

### Part 1.2 Create new dataframe team_regular_data

-   This dataframe contains each team's regular season data.
-   This dataframe features contain team, teamid, season, win_per, total_games, points, sd_points, rebs, fg_per, fg3_per, efg_per, ortg, opp_points, sd_opp_points, opp_rebs, opp_fg_per, opp_fg3_per, opp_efg_per, drtg, net_rtg.
-   opp_feature means the team opponent's feature value.
-   sd_points means the standard deviation of points.
-   Use function get_team_data(df, teamid, season)

```{r}
team_regular_data <- data.frame(team = character(0), teamid = numeric(0), season = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0))

teamid_lst <- unique(team_data$off_teamid)
season_lst <- unique(team_data$season)
df <- team_data[team_data$gametype == 2, ]


index <- 1
for (season in season_lst) {
  for (teamid in teamid_lst) {
    new_row <- get_team_data(df, teamid, season)
    team_regular_data[index, ] <- new_row
    index <- index + 1
  }
}

head(team_regular_data)
```

### Part 1.3 Create new dataframe sd_team_regular_data

-   This dataframe will contain the same columns as team_regular_data but with standardized values.
-   The league season average value will be 100, and the team new value will be (the original value / league average value) \* 100
-   For net_rtg, 0 is 100, then the standardized net_rtg will be 100 + original net_rtg value.
-   Use function standardize_team_data(df, teamid, season)

```{r}
sd_team_regular_data <- team_regular_data

for (i in 1: nrow(team_regular_data)) {
  teamid <- sd_team_regular_data[i, ]$teamid
  season <- sd_team_regular_data[i, ]$season
  new_row <- standardize_team_data(team_regular_data, teamid, season)
  sd_team_regular_data[i, ] <- new_row
}

head(sd_team_regular_data)
```

### Part 1.4 Create new dataframe playoff_series_data

-   This dataframe contains all playoff series result and two teams' regular season data.
-   This dataframe features include team1, team1id, team2, team2id, season, result, win_per, points, sd_points, rebs, fg_per, fg3_per, efg_per, ortg, opp_points, sd_opp_points, opp_rebs, opp_fg_per, opp_fg3_per, opp_efg_per, drtg, net_rtg, h2h.
-   team1 and team2 coould be swapped.
-   The feature result represents the result of the series, which is calculated by \# of team1 wins - \# of team2 wins. All the following features are the difference between team1 and team2, by team1 subtracting team2.
-   h2h is two teas regular season face to face result. It is computed by \# of team1 wins - \# of team2 wins.

```{r}
playoff_series_data <- data.frame(team1 = character(0), team1id = numeric(0), team2 = character(0), team2id = character(0), season = numeric(0), result = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0), h2h = numeric(0))

# create dataframe only with playoff games
df <- team_data[team_data$gametype == 4 & team_data$season >= 2014, ]

# add two new columns round and series
# use function get_round_series(gameid)
df <- df[order(df$gameid), ]
round_series <- lapply(df$gameid, get_round_series)
round_lst <- list()
series_lst <- list()
for (i in 1: length(round_series)) {
  round_lst <- append(round_lst, round_series[[i]][1])
  series_lst <- append(series_lst, round_series[[i]][2])
}

df$round <- unlist(round_lst)
df$series <- unlist(series_lst)

# get team1 and team2 info
# use function get_team1_team2(round, series)
round_lst <- c(1, 2, 3, 4)
series_lst <- c(8, 4, 2, 1)
index <- 1
for (season in unique(df$season)) {
  for (i in 1: 4) {
    round <- i
    for (series in 0: series_lst[i] - 1) {
      info <- get_team1_team2(df, season, round, series)
      info <- append(info, replicate(19, 0))
      playoff_series_data[index, ] <- unlist(info)
      index <- index + 1
    }
  }
}
playoff_series_data <- playoff_series_data[complete.cases(playoff_series_data), ]
rownames(playoff_series_data) <- NULL

# complete the dataframe
# the df should be sd_team_regular_data
season_lst <- playoff_series_data$season
team1id_lst <- playoff_series_data$team1id
team2id_lst <- playoff_series_data$team2id

for (i in 1: length(season_lst)) {
  new_row <- get_series_data(sd_team_regular_data, season_lst[i], team1id_lst[i], team2id_lst[i])
  series_result <- get_series_result(team_data, season_lst[i], team1id_lst[i], team2id_lst[i])
  h2h <- get_h2h(team_data, season_lst[i], team1id_lst[i], team2id_lst[i])
  new_row <- c(series_result, new_row, h2h)
  playoff_series_data[i, 6: ncol(playoff_series_data)] <- new_row
}

playoff_series_data <- playoff_series_data[, !names(playoff_series_data) %in% c("total_games")]
for (i in colnames(playoff_series_data)[6: ncol(playoff_series_data)]) {
  playoff_series_data[, i] <- as.numeric(playoff_series_data[, i])
}

head(playoff_series_data)
```

## Part 2. Data Visulization

### Part 2.1 points - series result graph

```{r}
plot(playoff_series_data$points, playoff_series_data$result, xlab = "points", ylab = "result", pch = 19, col = "lightblue")
```

### Part 2.2 sd_points - series result graph

```{r}
plot(playoff_series_data$sd_points, playoff_series_data$result, xlab = "sd_points", ylab = "result", pch = 19, col = "red")
```

### Part 2.3 opp_points - series result graph

```{r}
plot(playoff_series_data$opp_points, playoff_series_data$result, xlab = "opp_points", ylab = "result", pch = 19, col = "orange")
```

### Part 2.4 net_rtg - series result graph

```{r}
plot(playoff_series_data$net_rtg, playoff_series_data$result, xlab = "net_rtg", ylab = "result", pch = 19, col = "lightgreen")
```

### Part 2.5 efg - series result graph

```{r}
plot(playoff_series_data$efg_per, playoff_series_data$result, xlab = "efg", ylab = "result", pch = 19, col = "pink")
```

### Observation

From the above graphs, it is clear that efg and net_rtg has the most apparently linear relationship with series result.

## Part 3. Model Building

### Part 3.1 Create dataframe round1_2023

-   This dataframe should be similar to playoff_series_data. The only difference is there should be no result column in this dataframe.

```{r}
round1_2023 <- data.frame(team1 = character(0), team1id = numeric(0), team2 = character(0), team2id = character(0), season = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0), h2h = numeric(0))

team1_lst <- c("Boston Celtics", "New York Knicks", "Cleveland Cavaliers", "Milwaukee Bucks", "Oklahoma City Thunder", "Denver Nuggets", "Minnesota Timberwolves", "LA Clippers")
team2_lst <- c("Miami Heat", "Philadelphia 76ers", "Orlando Magic", "Indiana Pacers", "New Orleans Pelicans", "Los Angeles Lakers", "Phoenix Suns", "Dallas Mavericks")
season <- 2023

# use get_2023_series(df, team1, team2) to complete the dataframe
for (i in 1: 8) {
  new_row <- get_2023_series(sd_team_regular_data, team1_lst[i], team2_lst[i])
  h2h <- get_h2h(team_data, 2023, new_row[2], new_row[4])
  new_row <- c(new_row, h2h)
  round1_2023 [i, ] <- new_row
}

round1_2023 <- round1_2023[, !names(round1_2023) %in% c("total_games")]
for (i in colnames(round1_2023)[6: ncol(round1_2023)]) {
  round1_2023[, i] <- as.numeric(round1_2023[, i])
}

round1_2023
```

### Part 3.2 Build random forest model

-   This part I will build the random forest model.

```{r}
features <- colnames(round1_2023)[6: ncol(round1_2023)]
formula <- as.formula(paste0("result ~ ", paste(features, collapse = "+")))

# plot the error - number of trees figure
get_best_ntree(formula, playoff_series_data)
```

```{r}
# plot the importance plot
ntree <- 1000
get_importance(formula, playoff_series_data, best_mtry, ntree)
```

### Part 3.3 Round1 prediction

-   This part will predict round1 result and present teh predicted result.

```{r}
# build the model and predict the result
pred_result <- rep(0, 8)

for (i in 1: 10) {
  pred_result <- get_pred_result(formula, playoff_series_data, ntree, features, round1_2023) + pred_result
}

pred_result <- pred_result / 10
```

```{r}
# optimize the predicted result
round1_result <- optimize_result(pred_result)

round1_result
```

```{r}
# present the round1 predicted result
print_result(round1_2023, round1_result)
```

### Part 3.4 Round2 prediction

-   This part is for predicting round2 result based on previous predicted result

```{r}
# create round2_2023 dataframe
round2_2023 <- data.frame(team1 = character(0), team1id = numeric(0), team2 = character(0), team2id = character(0), season = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0), h2h = numeric(0))
team1_lst <- c("Boston Celtics", "New York Knicks", "Oklahoma City Thunder", "Denver Nuggets")
team2_lst <- c("Cleveland Cavaliers", "Milwaukee Bucks", "LA Clippers", "Phoenix Suns")

for (i in 1: 4) {
  new_row <- get_2023_series(sd_team_regular_data, team1_lst[i], team2_lst[i])
  h2h <- get_h2h(team_data, 2023, new_row[2], new_row[4])
  new_row <- c(new_row, h2h)
  round2_2023 [i, ] <- new_row
}

round2_2023 <- round2_2023[, !names(round2_2023) %in% c("total_games")]
for (i in colnames(round2_2023)[6: ncol(round2_2023)]) {
  round2_2023[, i] <- as.numeric(round2_2023[, i])
}

round2_2023
```

```{r}
# predict and present round2 predicted result
pred_result <- rep(0, 4)

for (i in 1: 10) {
  pred_result <- get_pred_result(formula, playoff_series_data, ntree, features, round2_2023) + pred_result
}

pred_result <- pred_result / 10
round2_result <- optimize_result(pred_result)

print_result(round2_2023, round2_result)
```

### Part 3.5 Round3 prediction

- This part is for predicting round3 result

```{r}
# create round3_2023 dataframe
round3_2023 <- data.frame(team1 = character(0), team1id = numeric(0), team2 = character(0), team2id = character(0), season = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0), h2h = numeric(0))
team1_lst <- c("Boston Celtics", "Oklahoma City Thunder")
team2_lst <- c("Milwaukee Bucks", "Denver Nuggets")

for (i in 1: 2) {
  new_row <- get_2023_series(sd_team_regular_data, team1_lst[i], team2_lst[i])
  h2h <- get_h2h(team_data, 2023, new_row[2], new_row[4])
  new_row <- c(new_row, h2h)
  round3_2023 [i, ] <- new_row
}

round3_2023 <- round3_2023[, !names(round3_2023) %in% c("total_games")]
for (i in colnames(round3_2023)[6: ncol(round3_2023)]) {
  round3_2023[, i] <- as.numeric(round3_2023[, i])
}

# predict and present result
pred_result <- rep(0, 2)

for (i in 1: 10) {
  pred_result <- get_pred_result(formula, playoff_series_data, ntree, features, round3_2023) + pred_result
}

pred_result <- pred_result / 10
round3_result <- optimize_result(pred_result)

print_result(round3_2023, round3_result)
```

### Part 3.6 NBA Final prediction

- This part is for prediction final result

```{r}
# create round4_2023 dataframe
round4_2023 <- data.frame(team1 = character(0), team1id = numeric(0), team2 = character(0), team2id = character(0), season = numeric(0), win_per = numeric(0), total_games = numeric(0), points = numeric(0), sd_points = numeric(0), rebs = numeric(0), fg_per = numeric(0), fg3_per = numeric(0), efg_per = numeric(0), ortg = numeric(0), opp_points = numeric(0), sd_opp_points = numeric(0), opp_rebs = numeric(0), opp_fg_per = numeric(0), opp_fg3_per = numeric(0), opp_efg_per = numeric(0), drtg = numeric(0), net_rtg = numeric(0), h2h = numeric(0))
team1_lst <- c("Boston Celtics")
team2_lst <- c("Oklahoma City Thunder")

for (i in 1: 1) {
  new_row <- get_2023_series(sd_team_regular_data, team1_lst[i], team2_lst[i])
  h2h <- get_h2h(team_data, 2023, new_row[2], new_row[4])
  new_row <- c(new_row, h2h)
  round4_2023 [i, ] <- new_row
}

round4_2023 <- round4_2023[, !names(round3_2023) %in% c("total_games")]
for (i in colnames(round4_2023)[6: ncol(round4_2023)]) {
  round4_2023[, i] <- as.numeric(round4_2023[, i])
}

# predict and present result
pred_result <- rep(0, 1)

for (i in 1: 10) {
  pred_result <- get_pred_result(formula, playoff_series_data, ntree, features, round4_2023) + pred_result
}

pred_result <- pred_result / 10
round4_result <- optimize_result(pred_result)

print_result(round4_2023, round4_result)
```

